### 2012-02-10 docs/SDC/sdc.tex
some updates

\maestro\ integrates the following system of equations:
\begin{eqnarray}
\frac{\partial\Ub}{\partial t} &=& 
    -\Ub\cdot\nabla\Ub  - \frac{1}{\rho}\nabla\pi 
    - \frac{\rho-\rho_0}{\rho} g\eb_r,\label{eq:momentum}\\
\frac{\partial(\rho X_k)}{\partial t} &=& 
    -\nabla\cdot(\rho X_k\Ub) + \rho\omegadot_k,\label{eq:species}\\
\frac{\partial(\rho h)}{\partial t} &=& 
    -\nabla\cdot(\rho h\Ub) + \frac{Dp_0}{Dt} 
    + \rho\Hnuc ,\label{eq:enthalpy}
\end{eqnarray}
together with a constraint equation.

By default, \maestro\ uses an operator-split approach to coupling
the different physical processes together (hydro, reactions, diffusion).
With operator-splitting, we
react for a half timestep, then perform the hydrodynamics and thermal
diffusion on this intermediate state, and finally react for the
remaining half timestep to advance the solution.  For problems where
the reactions greatly alter the energy balance or composition, this
operator splitting approach can lead to inaccuracies in the solution.
This can be more of an issue for low Mach number methods that can take
large hydrodynamic time steps.

An alternate approach to coupling is spectral deferred corrections
(SDC). SDC is an iterative approach to couple the various operations
together, with each operator seeing an iteratively lagged
approximation of the other operations as a source term.  The SDC
algorithm converges to an integral representation of the solution in
time that couples all of the processes together in a self-consistent
fashion, see \cite{Non11}.

As a first attempt, we will work on coupling hydro and reactions only
via SDC, with a time-independent base state.

\section{Strang-split Overview}

In the Strang-split version of \maestro, the reaction and advection
processes operate independent of one-another.  The species and
enthalpy equations are updated through $\Delta t$ using the following
sequence:

@@

\item React for $\Delta t/2$

In the Strang-split version of \maestro, the reaction network solves just
the reaction portion of the species evolution equations along with a
temperature evolution equation.  This is done using a standard ODE solver
(like {\tt VODE}) on the system:
\begin{eqnarray}
\frac{dX_k}{dt} &=& \omegadot_k(\rho,X_k,T) \\
\frac{dT}{dt}   &=& 
    \frac{1}{c_p} \left ( -\sum_k \xi_k  \omegadot_k  + \Hnuc \right ).
\end{eqnarray}
Here, $T$ is evolved solely to evaluate the reaction rates,
$\omegadot_k(\rho,X_k,T)$.  Furthermore, we simplify the problem
``freezing'' the thermodynamics---i.e., $c_p$ and $\xi_k$ are evaluated at the
start of the integration and held constant thereafter.

We update the enthalpy at the end of the integration as:
\begin{equation}
h^\mathrm{new} = h^\mathrm{old} + \frac{\Delta t}{2} \Hnuc
\end{equation}

\item Advect for $\Delta t$

Here we solve
\begin{eqnarray}
\frac{\partial(\rho X_k)}{\partial t} &=& 
    -\nabla\cdot(\rho X_k\Ub) \\
\frac{\partial(\rho h)}{\partial t} &=& 
    -\nabla\cdot(\rho h\Ub) + \frac{Dp_0}{Dt} 
\end{eqnarray}

Note that no explicit reaction terms appear here.  Since the advection
takes place using the state updated from the reaction step, the effect
of the reactions is implicitly contained in the advective update.

\item React for $\Delta t/2$

@@

\section{SDC Without Thermal Diffusion or Base State Evolution}
In the SDC version, the VODE integration at the end of an SDC
iteration is responsible for updating all the thermodynamic quantities
\{$\rho$, $(\rho h)$, and $(\rho X_k$)\}, including both the advective
fluxes (incorporated via a piecewise constant advective flux source
term) and the reactions.  This provides a much stronger coupling between
the physical processes.  In particular, our system now looks like:
\begin{eqnarray}
\frac{d(\rho X_k)}{dt} &=& \rho \omegadot_k(\rho,X_k,T) + A_{\rho X_k} \label{eq:sdc:rhoX} \\
\frac{d(\rho h)}{dt}   &=& \rho \Hnuc + A_{\rho h} \label{eq:sdc:rhoh}
\end{eqnarray}
Here, $A_{\rho X_k}$ and $A_{\rho h}$ are piecewise-constant (in time)
approximations to the change in ${\rho X_k}$ and ${\rho h}$ (respectively)
due to the advection.  These are constructed by calling {\tt density\_advance}
and {\tt enthalpy\_advance} in \maestro\ and passed into the network solver
during the reaction step.  A flowchart of the \maestro\ SDC algorithm is 
shown in figure~\ref{fig:sdc:flowchart}.

@@

In the advective update, our goal is to compute $A_{\rho X_k}$ and
$A_{\rho h}$.  These terms approximate the following:
\begin{eqnarray}
A_{\rho X_k} &=&  \left [- \nabla \cdot (\rho X_k \Ub) \right ]^{n+1/2} \\
A_{\rho h}   &=&  \left [- \nabla \cdot (\rho h U) + \frac{Dp_0}{Dt} \right ]^{n+1/2}
\end{eqnarray}
The construction of the interface states used in the advective terms
uses an iteratively-lagged approximation to the reaction source terms
($I_{\rho X_k}$ and $I_{\rho h}$, see below) in the interface
prediction.  This explicitly couples the reaction process to the
advection process.

@@

The RHS routine that the ODE solver operates on will first construct
the density as:
\begin{equation}
\rho = \sum_k (\rho X_k)
\end{equation}
It will then derive the temperature from the equation of state.  If we
are running with {\tt use\_tfromp = T}, then we do
\begin{equation}
T = T(\rho, p_0, X_k)
\end{equation}
otherwise, we do
\begin{equation}
T = T(\rho, h, X_k)
\end{equation}
Note that in constrast to the Strang-split version, here we call the EOS
every time we enter the RHS routine.

Finally we integrate the ODE system (Eqs.~\ref{eq:sdc:rhoX} and \ref{eq:sdc:rhoh}).
At the end of the integration, we define $I_{\rho X_k}$ and $I_{\rho
  h}$.  The actual form of these depends on what we are predicting.
We note that we only need $I_{\rho X_k}$ and $I_{\rho h}$ for the
prediction of the interface states, not the final conservative update.
That is because all we need from the advection solver is the
approximation to $A_{\rho X_k}$ and $A_{\rho h}$, not the final
updated state.

@@

Note also that the SDC integration by {\tt VODE} does not operate on 
the velocities at all.  That update is handled in the same fashion 
as the Strang-split version of the code.

@@

\section{Algorithm Flowchart - ADR with Fixed Base State}
We now include thermal diffusion and assume the base state is constant in time but not space:
\begin{eqnarray}
\frac{\partial\Ub}{\partial t} &=& 
    -\Ub\cdot\nabla\Ub  - \frac{1}{\rho}\nabla\pi 
    - \frac{\rho-\rho_0}{\rho} g\eb_r,\\
\frac{\partial(\rho X_k)}{\partial t} &=& 
    -\nabla\cdot(\rho X_k\Ub) + \rho\omegadot_k,\label{eq:sdc species}\\
\frac{\partial(\rho h)}{\partial t} &=&
    -\nabla\cdot(\rho h\Ub) + \underbrace{\Ub\cdot\nabla p_0}_{Dp_0/Dt}
    + \rho\Hnuc
    + \underbrace{\nabla\cdot\frac{\kth}{c_p}\nabla h - \sum_k\nabla\cdot\frac{\xi_k k_{\rm th}}{c_p}\nabla X_k - \nabla\cdot\frac{h_p k_{\rm th}}{c_p}\nabla p_0}_{\nabla\cdot k_{\rm th}\nabla T}.\label{eq:sdc enthalpy}
\end{eqnarray}

@@

{\bf Step 1D:} ({\it Advance thermodynamic variables}) Define $Q_{\rho X_k}^{\rm pred}$ as the right hand side of (\ref{eq:sdc species 2}) without the $I_{R,\rho X_k}^{\rm lagged}$ term, and define $Q_{\rho h}^{\rm pred}$ as the right hand side of (\ref{eq:sdc enthalpy 2}) without the $I_{R,\rho h}^{\rm lagged}$ term.  Use VODE to integerate (\ref{eq:sdc species}) and (\ref{eq:sdc enthalpy}) over $\Delta t$ to advance $(\rho X_k, \rho h)^n$ to $(\rho X_k, \rho h)^{n+1,{\rm pred}}$ using the piecewise-constant advection and diffusion source terms:
\begin{eqnarray}
\frac{\partial(\rho X_k)}{\partial t} &=& Q_{\rho X_k}^{\rm pred} + \rho\dot\omega_k\\
\frac{\partial(\rho h)}{\partial t} &=& Q_{\rho h}^{\rm pred} + \rho\Hnuc.
\end{eqnarray}

@@

{\bf Step 2D:} ({\it Advance thermodynamic variables}) Define $Q_{\rho X_k}$ as the right hand side of (\ref{eq:sdc species 3}) without the $I_{R,\rho X_k}^{\rm pred}$ term, and define $Q_{\rho h}$ as the right hand side of (\ref{eq:sdc enthalpy 3}) without the $I_{R,\rho h}^{\rm pred}$ term.  Use VODE to integerate (\ref{eq:sdc species}) and (\ref{eq:sdc enthalpy}) over $\Delta t$ to advance $(\rho X_k, \rho h)^n$ to $(\rho X_k, \rho h)^{n+1}$ using the piecewise-constant advection and diffusion source terms:
\begin{eqnarray}
\frac{\partial(\rho X_k)}{\partial t} &=& Q_{\rho X_k} + \rho\dot\omega_k\\
\frac{\partial(\rho h)}{\partial t} &=& Q_{\rho h} + \rho\Hnuc.
\end{eqnarray}

>>>>

Spectral deferred corrections (SDC) is an iterative scheme used to integrate
the thermodynamic variables, $(\rho X_k,\rho h)$, over a time step.  It has
been shown to be more accurate and efficient than Strang splitting in a
terrestrial, non-stratified, low Mach number reacting flow solver \cite{Non11},
so we would like to develop an SDC version of \maestro.

\maestro\ integrates the following system of equations:
\begin{eqnarray}
\frac{\partial\Ub}{\partial t} &=& 
    -\Ub\cdot\nabla\Ub  - \frac{1}{\rho}\nabla\pi 
    - \frac{\rho-\rho_0}{\rho} g\eb_r,\label{eq:momentum}\\
\frac{\partial(\rho X_k)}{\partial t} &=& 
    -\nabla\cdot(\rho X_k\Ub) + \rho\omegadot_k,\label{eq:species}\\
\frac{\partial(\rho h)}{\partial t} &=& 
    -\nabla\cdot(\rho h\Ub) + \frac{Dp_0}{Dt} 
    + \rho\Hnuc + \nabla\cdot k_{\rm th}\nabla T,\label{eq:enthalpy}
\end{eqnarray}
together with base state evolution equations and a constraint equation.
By default, \maestro\ advances the thermodynamic variables by coupling
the different physical processes together (advection, diffusion, reactions) using
Strang splitting.  Specifically, we integrate the reaction terms 
over half a time step ignoring contributions from advection and diffusion,
then from this intermediate state we integrate the advection and diffusion terms over
a full time step (while ignoring reactions), and finally
integrate the reactions over a half time step (ignoring advection and diffusion).
For problems where
the reactions and/or diffusion greatly alter the energy balance or composition 
as compared to advection, this operator splitting approach can lead to large 
splitting errors and highly inaccurate solutions.
This issue is can be particularly exasperating 
for low Mach number methods that can take large advection-based time steps.

An alternate approach to advancing the thermodynamic variables is SDC. 
SDC is an iterative approach to couple the various processes
together, with each process seeing an 
approximation of the other processes as a source term.  The SDC
algorithm converges to an integral representation of the solution in
time that couples all of the processes together in a self-consistent
fashion, see \cite{Non11}.

As a first attempt, we will work on coupling advection and reactions only
via SDC, with a base state that is fixed in time, but not space.

\section{Strang-Splitting Without Thermal Diffusion or Base State Evolution}

In the Strang splitting version of \maestro, the reaction and advection
processes operate independent of one-another.  The species and
enthalpy equations are integrated over $\Delta t$ using the following
sequence:

@@

\item React for $\Delta t/2$:

In the Strang splitting version of \maestro, the reaction network solves just
the reaction portion of the species evolution equations along with a
temperature evolution equation.  This is done using a standard stiff ODE solver
package (like {\tt VODE}) on the system:
\begin{eqnarray}
\frac{dX_k}{dt} &=& \omegadot_k(\rho,X_k,T), \\
\frac{dT}{dt}   &=& 
    \frac{1}{c_p} \left ( -\sum_k \xi_k  \omegadot_k  + \Hnuc \right ).
\end{eqnarray}
Here, $T$ is evolved solely to evaluate the reaction rates,
$\omegadot_k(\rho,X_k,T)$.  Furthermore, we simplify the problem
``freezing'' the thermodynamics---i.e., $c_p$ and $\xi_k$ are evaluated at the
start of the integration and held constant thereafter.
The density remains constant during this step, i.e., 
$\rho^\mathrm{new} = \rho^\mathrm{old}$, and we
update the enthalpy at the end of the integration as:
\begin{equation}
h^\mathrm{new} = h^\mathrm{old} + \frac{\Delta t}{2} \Hnuc.
\end{equation}

\item Advect for $\Delta t$:

Beginning with the results from the previous reaction half time step, we integrate
that state using the equations
\begin{eqnarray}
\frac{\partial(\rho X_k)}{\partial t} &=& 
    -\nabla\cdot(\rho X_k\Ub), \\
\frac{\partial(\rho h)}{\partial t} &=& 
    -\nabla\cdot(\rho h\Ub) + \frac{Dp_0}{Dt}.
\end{eqnarray}
Note that no reaction terms appear here.  Since the advection
takes place using the state updated from the reaction step, the effect
of the reactions is implicitly contained in the advective update.

\item React for $\Delta t/2$:

@@

Note that \maestro\ uses a predictor-corrector approach.  After integrating $(\rho X_k,\rho h)$ over
the time step, we use this time-advanced state to get a better estimate of a time-centered $\beta_0$
and $S$.  We re-compute the advective velocities using an updated divergence constraint and repeat 
the thermodynamic variable advance.

\section{SDC Without Thermal Diffusion or Base State Evolution}
In the SDC version, the {\tt VODE} integration at the end of an SDC
iteration is responsible for updating all the thermodynamic quantities
including both the advection (incorporated via a piecewise constant advective 
flux divergence source term) and the reactions.  This provides a much stronger coupling between
the physical processes.  In particular, our system now looks like:
\begin{eqnarray}
\frac{d(\rho X_k)}{dt} &=& \rho \omegadot_k(\rho,X_k,T) + \underbrace{\nabla\cdot(\rho X_k\Ub)}_{A_{\rho X_k}}\label{eq:sdc:rhoX} \\
\frac{d(\rho h)}{dt}   &=& \rho \Hnuc + \underbrace{\nabla\cdot(\rho h\Ub)}_{A_{\rho h}} \label{eq:sdc:rhoh}
\end{eqnarray}
Here, $A_{\rho X_k}$ and $A_{\rho h}$ are piecewise-constant (in time)
approximations to the change in ${\rho X_k}$ and ${\rho h}$ (respectively)
due to the advection.  These are constructed by calling {\tt density\_advance}
and {\tt enthalpy\_advance} in \maestro\ and passed into the network solver
during the reaction step.  A flowchart of the \maestro\ SDC algorithm is 
shown in figure~\ref{fig:sdc:flowchart}.

@@

In the advective update, our goal is to compute $A_{\rho X_k}$ and
$A_{\rho h}$.  These terms approximate the following:
\begin{eqnarray}
A_{\rho X_k} &=&  \left [- \nabla \cdot (\rho X_k \Ub) \right ]^{n+1/2} \\
A_{\rho h}   &=&  \left [- \nabla \cdot (\rho h \Ub) + \frac{Dp_0}{Dt} \right ]^{n+1/2}
\end{eqnarray}
The construction of the interface states used in the advective terms
uses either a time-lagged or iteratively-lagged approximation to the reaction
terms ($I_{\rho X_k}$ and $I_{\rho h}$, see below) as a source term in the interface
prediction.  This explicitly couples the reaction process to the
advection process.

@@

The RHS routine that the ODE solver operates on will first construct
the density as:
\begin{equation}
\rho = \sum_k (\rho X_k)
\end{equation}
It will then derive the temperature from the equation of state.  If we
are running with {\tt use\_tfromp = T}, then we do
\begin{equation}
T = T(\rho, p_0, X_k)
\end{equation}
otherwise, we do
\begin{equation}
T = T(\rho, h, X_k)
\end{equation}
Note that in contrast to the Strang splitting version, here we call the EOS
every time we enter the RHS routine, but here we call the EOS to compute temperature
rather than thermodynamic coefficients.

Finally we integrate the ODE system (Eqs.~\ref{eq:sdc:rhoX} and \ref{eq:sdc:rhoh}).
At the end of the integration, we define $I_{\rho X_k}$ and $I_{\rho h}$.  The actual 
form of these depends on what quantities we predict to edges during 
the construction of the advective fluxes.
Note that we only need $I_{\rho X_k}$ and $I_{\rho h}$ for the
prediction of the interface states, and not the {\tt VODE} integration.
This is because all we need from the advection solver is the
approximation to $A_{\rho X_k}$ and $A_{\rho h}$ and not the final
updated state.

@@

Note also that the SDC integration by {\tt VODE} does not operate on 
the velocities at all.  That update is handled in the same fashion 
as the Strang splitting version of the code.

@@

\section{Algorithm Flowchart - ADR with Fixed Base State}
We now include thermal diffusion and assume the base state is constant in time but not space:
\begin{eqnarray}
\frac{\partial\Ub}{\partial t} &=& 
    -\Ub\cdot\nabla\Ub  - \frac{1}{\rho}\nabla\pi 
    - \frac{\rho-\rho_0}{\rho} g\eb_r,\\
\frac{\partial(\rho X_k)}{\partial t} &=& 
    -\nabla\cdot(\rho X_k\Ub) + \rho\omegadot_k,\label{eq:sdc species}\\
\frac{\partial(\rho h)}{\partial t} &=&
    -\nabla\cdot(\rho h\Ub) + \underbrace{\Ub\cdot\nabla p_0}_{Dp_0/Dt}
    + \rho\Hnuc
    + \underbrace{\nabla\cdot\frac{\kth}{c_p}\nabla h - \sum_k\nabla\cdot\frac{\xi_k k_{\rm th}}{c_p}\nabla X_k - \nabla\cdot\frac{h_p k_{\rm th}}{c_p}\nabla p_0}_{\nabla\cdot k_{\rm th}\nabla T}.\nonumber\\
\label{eq:sdc enthalpy}
\end{eqnarray}

@@

{\bf Step 1D:} ({\it Advance thermodynamic variables}) Define $Q_{\rho X_k}^{\rm pred}$ as the right hand side of (\ref{eq:sdc species 2}) without the $I_{R,\rho X_k}^{\rm lagged}$ term, and define $Q_{\rho h}^{\rm pred}$ as the right hand side of (\ref{eq:sdc enthalpy 2}) without the $I_{R,\rho h}^{\rm lagged}$ term.  Use {\tt VODE} to integerate (\ref{eq:sdc species}) and (\ref{eq:sdc enthalpy}) over $\Delta t$ to advance $(\rho X_k, \rho h)^n$ to $(\rho X_k, \rho h)^{n+1,{\rm pred}}$ using the piecewise-constant advection and diffusion source terms:
\begin{eqnarray}
\frac{\partial(\rho X_k)}{\partial t} &=& Q_{\rho X_k}^{\rm pred} + \rho\dot\omega_k\\
\frac{\partial(\rho h)}{\partial t} &=& Q_{\rho h}^{\rm pred} + \rho\Hnuc.
\end{eqnarray}

@@

{\bf Step 2D:} ({\it Advance thermodynamic variables}) Define $Q_{\rho X_k}$ as the right hand side of (\ref{eq:sdc species 3}) without the $I_{R,\rho X_k}^{\rm pred}$ term, and define $Q_{\rho h}$ as the right hand side of (\ref{eq:sdc enthalpy 3}) without the $I_{R,\rho h}^{\rm pred}$ term.  Use {\tt VODE} to integerate (\ref{eq:sdc species}) and (\ref{eq:sdc enthalpy}) over $\Delta t$ to advance $(\rho X_k, \rho h)^n$ to $(\rho X_k, \rho h)^{n+1}$ using the piecewise-constant advection and diffusion source terms:
\begin{eqnarray}
\frac{\partial(\rho X_k)}{\partial t} &=& Q_{\rho X_k} + \rho\dot\omega_k\\
\frac{\partial(\rho h)}{\partial t} &=& Q_{\rho h} + \rho\Hnuc.
\end{eqnarray}
